\documentclass[conference]{IEEEtran}

\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{hyperref}
\usepackage{siunitx}

\graphicspath{{figures/}}

\begin{document}

\title{Energy Consumption of Large Language Model Inference:\\A Multi-Architecture Benchmarking Study on GPU Hardware}

\author{\IEEEauthorblockN{Daniel Cregg}
\IEEEauthorblockA{South East Technological University\\
Waterford, Ireland\\
daniel.cregg@setu.ie}}

\maketitle

\begin{abstract}
We present a rigorous framework for measuring the energy consumption of large language model (LLM) inference on GPU hardware and apply it to benchmark 13 open-weight models spanning 1B to 70B parameters across five architectural families (Llama, Mistral, Qwen, Phi, Gemma). Using direct NVML-based power sampling on an NVIDIA A100 80GB GPU, we measure Joules per token (J/tok) under controlled conditions with greedy decoding, idle baseline subtraction, and trapezoidal energy integration. Our results reveal a sub-linear scaling law relating energy to model size across architectures, confirm the bitsandbytes quantisation energy paradox previously observed on a single model family, and quantify the batch size efficiency frontier. We additionally ground our measurements in Irish electricity grid carbon intensity data, demonstrating a 2.9$\times$ variation in per-token carbon emissions depending on time of day. All code, data, and reproducibility artifacts are publicly available.
\end{abstract}

\begin{IEEEkeywords}
energy efficiency, large language models, GPU benchmarking, inference energy, carbon emissions, sustainable AI
\end{IEEEkeywords}

\section{Introduction}

The rapid deployment of large language models (LLMs) has raised urgent questions about their energy footprint. While training costs have received significant attention~\cite{strubell2019energy, patterson2021carbon}, inference increasingly dominates total lifecycle energy as models are deployed at scale~\cite{desislavov2023trends}. Understanding and optimising inference energy is therefore critical for sustainable AI deployment.

Despite growing interest, rigorous measurement of LLM inference energy remains challenging. Software-based estimators (e.g., CodeCarbon) rely on simplified models that may not accurately reflect GPU power dynamics~\cite{jay2023experimental}. Hardware-level measurement via NVIDIA's Management Library (NVML) provides direct power readings but requires careful instrumentation, including SLURM GPU mapping, idle baseline subtraction, and appropriate integration methods.

This paper makes the following contributions:
\begin{enumerate}
    \item A validated, open-source framework for measuring LLM inference energy using direct NVML power sampling with trapezoidal integration and idle baseline subtraction.
    \item A comprehensive benchmark of 13 open-weight models (1B--70B parameters) across five architectural families on an A100 80GB GPU.
    \item Empirical scaling laws relating energy to model size across architectures, extending prior work limited to the Pythia family.
    \item Quantification of the bitsandbytes quantisation energy paradox on Llama-3.1-8B with FP16, INT8, and INT4 precisions.
    \item Carbon impact analysis grounded in real Irish electricity grid data, demonstrating temporal variation in per-token emissions.
\end{enumerate}

\section{Background and Related Work}

\subsection{LLM Energy Measurement}

Prior work on measuring LLM energy falls into three categories: software estimators~\cite{lacoste2019quantifying, lannelongue2021green}, hardware-level measurement~\cite{samsi2023words, argerich2024measuring}, and analytical models~\cite{desislavov2023trends}. Software estimators like CodeCarbon use thermal design power (TDP) scaling, which introduces systematic error~\cite{jay2023experimental}. Hardware-level approaches using NVML or power meters provide higher accuracy but require careful experimental design.

Samsi et al.~\cite{samsi2023words} measured inference energy for several LLMs using NVML on an A100 cluster, finding sub-linear scaling with model size. Argerich and Pati\~no-Mart\'inez~\cite{argerich2024measuring} extended this to include quantised models, observing that na\"ive quantisation can increase total energy despite reducing power draw.

\subsection{The Quantisation Energy Paradox}

On-the-fly dequantisation via bitsandbytes incurs a throughput penalty that can dominate the power reduction from lower-precision arithmetic. Prior work on the energy-bench framework found that INT8 quantisation of Mistral-7B via bitsandbytes increased J/tok by 2.88$\times$ despite a 37\% reduction in power draw, because throughput dropped 4.5$\times$. Pre-quantised kernels (AWQ, GPTQ via vLLM) did not exhibit this paradox.

\subsection{Carbon-Aware Computing}

The carbon impact of computation depends on both energy consumed and the carbon intensity of the electricity grid. Dodge et al.~\cite{dodge2022measuring} demonstrated that scheduling ML training to align with low-carbon grid periods can significantly reduce emissions. We extend this concept to inference, using real Irish grid data from EirGrid.

\section{Framework Design}

Our framework follows a four-layer architecture:

\begin{enumerate}
    \item \textbf{Hardware Instrumentation} (Layer 1): NVML-based power sampling in a background thread at 10\,Hz, with SLURM GPU index resolution and trapezoidal energy integration.
    \item \textbf{Inference Engine} (Layer 2): Standardised model loading (FP16, INT8, INT4 via bitsandbytes) and inference execution with \texttt{torch.cuda.synchronize()} for precise timing.
    \item \textbf{Metric Computation} (Layer 3): Computes J/tok, tok/s, mean/peak watts, and optional gCO$_2$eq/tok from energy and inference measurements.
    \item \textbf{Orchestration} (Layer 4): Runs the benchmark protocol (3 warmup + 10 measured runs per configuration) and produces structured JSON reports with full metadata.
\end{enumerate}

Key design decisions include: (1) NVML-only power measurement (no software estimators), (2) idle baseline subtraction for net inference energy, (3) greedy decoding only for deterministic output, (4) 10 measured runs with 95\% confidence intervals, and (5) a fixed 5-prompt set spanning summarisation, QA, code generation, long-form explanation, and reasoning tasks.

\section{Implementation}

% TODO: Fill in implementation details from actual runs

\subsection{Hardware}

All experiments were conducted on an NVIDIA A100 80GB PCIe GPU accessed via SLURM. The A100 has a TDP of 300\,W and an observed idle power of approximately 40--50\,W.

\subsection{Software}

The framework is implemented in Python using PyTorch, Hugging Face Transformers, and pynvml. Quantisation uses bitsandbytes for INT8 and INT4 (NF4) precision.

\section{Validation}

% TODO: Fill in validation results from Llama-3.2-1B

Before the full benchmark campaign, we validated the framework on \texttt{meta-llama/Llama-3.2-1B-Instruct}. Validation criteria included: J/tok in the expected range [0.01, 0.2] for a 1B model, idle baseline between 30--60\,W, and coefficient of variation (CV) below 15\% across 10 runs.

\section{Empirical Benchmark Study}

% TODO: Fill in results from all 13 models

\subsection{Model Selection}

We benchmark 13 models spanning five architectural families and parameter counts from 1B to 70B (Table~\ref{tab:summary}).

\subsection{Scaling Law Results}

% TODO: Reference Figure 1 with fitted exponent

\subsection{Batch Size Effects}

% TODO: Reference Figure 3

\subsection{Quantisation Impact}

% TODO: Reference Figure 4 with multiplier comparison

\subsection{Architecture Comparison}

% TODO: Reference Figure 6

\subsection{Carbon Impact}

% TODO: Reference Figure 7 with Irish grid data

\section{Discussion}

\subsection{Comparison with Prior Work}

% TODO: Fill in comparison with energy-bench results

\subsection{Limitations}

Our study has several limitations: (1) single GPU type (A100 PCIe), (2) bitsandbytes-only quantisation (not AWQ/GPTQ), (3) greedy decoding only, (4) English-only prompts, and (5) inference only (not training).

\subsection{Implications}

% TODO: Fill in practical implications

\section{Conclusion and Future Work}

% TODO: Fill in conclusions

\bibliographystyle{IEEEtran}
\bibliography{references}

\end{document}
